Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime where a formula like~\eqref{eq:MAP_quadratic} is valid.

We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
This is true for
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.
This is true for categorical or Dirichlet distributions.

With some more work, it is possible to get a wealth of results under this assumption.
For instance \citet{ostrovskii2021finite} characterizes the number of samples needed to reach this quadratic behavior for arbitrary parametric models, under the assumptions of self-concordance, or pseudo self-concordance \citep{bach2010self}.
Or in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to quantify when $\logpart$ can be sandwiched between quadratics in $\nat$, but this does not directly translate into quadratics in $\m$.

All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.

\citet{anastasiou2017bounds} and
\citet{marteauferey2019beyond})
\fdk{
We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate.
[Robust stochastic approximation approach to stochastic programming]. (The paper uses a bounded stochastic gradient assumption, but can be relaxed to bounded variance instead.) However, strong-convexity in the Euclidean norm is not strictly necessary to find an algorithm with convergence rate of $O(1/n)$, as shown by [Non-strongly-convex smooth stochastic approximation with convergence rate O(1/n)] for linear and logistic regression. Their proof techniques based on self-concordance have been extended to other setting, [Finite-sample analysis of M-estimators using self-concordance, Learning Exponential Families in High-Dimensions: Strong Convexity and Sparsity].
The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
}

>>>>

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

The entropy $\conj$ is self-concordant for a lot of common families.
For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
as is the case with:
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g. categorical and Dirichlet distributions.

There are quite a few pieces of work in this direction. 
\citet{ostrovskii2021finite} characterizes the number of samples needed to reach the asymptotic quadratic behavior, for any parametric models with a self-concordant log-likelihood.
\citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
More closely, in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to obtain a similar result.
However these results are expressed with quadratics in $\nat$, not $\m$, and they do not directly translate to convergence rates for the MAP, but they might with some more work.

These results partially answer~\eqref{problem}. 
Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us investigate a bias-variance decomposition to formulate a conjecture.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
